---
title: "Descriptive Graphics in R"
author: "Brandon M. Greenwell"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required data
eggshell <- read.csv("C:\\Users\\greenweb\\Desktop\\Filing cabinet\\STT 6300\\Data sets\\eggshell.csv",
                     header = TRUE)
```

## Always plot your data!!

* Descriptive statistics are useful, but should always be complimented with appropriate graphics
* We want to see what the distribution of the sample looks like
* The "shape" of the distribution can often tell us a lot about the population of interest
* A few common graphics we will cover here:
    - boxplot
    - histogram
    - smoothed histogram/nonparametric density
    

## Boxplots

* A boxplot is obtained by simply drawing a box!
* The top/bottom of the box correspond to the upper/lower quartiles
* A line is then drawn through the median
* Sometimes, the sample mean is also displayed (depends on the software)
* Vertical lines are then drawn from the top and bottom of the box (these are called whiskers)
* These vertical lines extend to the most extreme values in the sample within the range
$$
Q_1 - 1.5 \times IQR \quad \textrm{and} \quad Q_3 + 1.5 \times IQR
$$
* Observations outside this range are denoted by individual points and are considered as potential outliers!


## Boxplots

```{r, eval=FALSE}
# Set up plotting region
par(mfrow = c(1, 2))

# Boxplot for thickness
boxplot(eggshell$thickness, ylab = "Thickness (mm)", col = "red")

# Boxplot for PCB concentration
boxplot(eggshell$pcb, ylab = "PCB concentration (ppb)", col = "blue")
```

## Boxplots

```{r, echo=FALSE}
# Set up plotting region
par(mfrow = c(1, 2))

# Boxplot for thickness
boxplot(eggshell$thickness, ylab = "Thickness (mm)", col = "red")

# Boxplot for PCB concentration
boxplot(eggshell$pcb, ylab = "PCB concentration (ppb)", col = "blue")
```


## Histograms

* Another useful graphic for displaying the shape of a distribution is the histogram
* The idea is to divide the observations into bins of equal width
* We then draw rectangles for each bin starting at zero and with a height that corresponds to either the frequency (i.e., the number of observations) in the bin or the relative frequency (i.e., the percentage of observations)


## Histograms

```{r, eval=FALSE}
# Set up plotting region
par(mfrow = c(1, 3))

# Default histogram for thickness
hist(eggshell$thickness, main = "", xlab = "Thickness (mm)")

# Too many bins
hist(eggshell$thickness, breaks = 25, freq = FALSE, main = "",
     xlab = "Thickness (mm)")

# Too few bins
hist(eggshell$thickness, breaks = 2, freq = FALSE, main = "",
     xlab = "Thickness (mm)")
```


## Histograms

```{r, echo=FALSE}
# Set up plotting region
par(mfrow = c(1, 3))

# Default histogram for thickness
hist(eggshell$thickness, main = "", xlab = "Thickness (mm)")

# Too many bins
hist(eggshell$thickness, breaks = 25, freq = FALSE, main = "",
     xlab = "Thickness (mm)")

# Too few bins
hist(eggshell$thickness, breaks = 2, freq = FALSE, main = "",
     xlab = "Thickness (mm)")
```


## Distributional shapes

* Determining the number, or equivalently, the width of the bins can be difficult
* If there are too many bins, there will not be enough (if any) in each group
* If there are too few bins, then it will be difficult to capture the shape of the distribution
* If we had an infinite amount of data we could make the histogram as fine as possible by increasing the number of bins
* In this limiting case, the histogram will look like a smooth curve, which we call a density curve


## Distributional shapes

```{r, echo=FALSE, fig.wdith=5, fig.height=5}
set.seed(101)
x <- seq(from = 0, to = 4, length = 500)
par(mfrow = c(2, 2))
plot(dnorm, xlim = c(-4, 4), xlab = "", ylab = "Density")
plot(dlnorm, xlim = c(0, 4), xlab = "", ylab = "Density")
plot(x, rev(dlnorm(x)), type = "l", xlab = "", ylab = "Density")
plot(density(c(rnorm(5000), rnorm(5000, mean = 4, sd = 1.5))), 
     main = "", xlab = "", ylab = "Density")
```
