---
title: "Descriptive Graphics in R"
author: "Brandon M. Greenwell"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required data
eggshell <- read.csv("C:\\Users\\greenweb\\Desktop\\Filing cabinet\\STT 6300\\Data sets\\eggshell.csv",
                     header = TRUE)
```

## Always plot your data!!

* Descriptive statistics are usefull, but should always be complimented with appropriate graphics
* We want to see what the distribution of the sample looks like
* The "shape" of the distribution can often tell us a lot about the population of interest
* A few common graphics we will covere here:
    - boxplot
    - histogram
    - smoothed histogram/nonparametric density
    

## Boxplots

* A boxplot is obtained by simply drawing a box!
* The top/bottom of the box correspond to the upper/lower quartiles
* A line is then drawn through the median
* Sometimes, the sample mean is also displayed (depends on the software)
* Vertical lines are then drawn from the top and bottom of the box (these are called whiskers)
* These vertical lines extend to the most extreme values in the sample within the range
$$
Q_1 - 1.5 \times IQR \quad \textrm{and} \quad Q_3 + 1.5 \times IQR
$$
* Observations outside this range are denoted by individual points and are considered as potential outliers!


## Boxplots

```{r}
par(mfrow = c(1, 2))
boxplot(eggshell$thickness, ylab = "Thickness (mm)", col = "red")
boxplot(eggshell$pcb, ylab = "PCB concentration (ppb)", col = "blue")
```
